### 2018-05-27 write-up/ott_spec.tex
Finish Conclusion

The following pages present a specification of the grammar and type-system used
by my project, produced using the Ott\cite{sewell_ott} tool. It is important to note
that the type-system described here is not how it is implemented: it is easier
and clearer to describe the system as below for explaining. However, for implementing,
I found it much more and user- and debugging-friendly to:

@@

    \item Implement it so that the type-environment \emph{changes} as a result
        of type-checking an expression; with this, the below semantics describe
        the \emph{difference} between the environment after and before checking
        an expression.

@@

    \item Have \emph{one} environment where variables were \emph{tagged} as
        linear and un-used, linear and used, and intuitionistic. This was
        definitely an implementation convenience so that variable binding could
        be handled uniformly for linear and intuitionistic variables and
        scoping/variable look-up could be handled implicitly thanks to the
        associative-list structure of the environment.

>>>>

The following pages present a specification of the grammar and type-system used
by my project, produced using the Ott~\cite{sewell_ott} tool. It is important to note
that the type-system described here is not how it is implemented: it is easier
and clearer to describe the system as below for explaining. However, for implementing,
I found it much more and user- and debugging-friendly to:

@@

    \item Implement it so that the type-environment \emph{changes} as a result
        of type-checking an expression, similar to the rules shown in
        Figure~\ref{fig:example_rules}; with this, the below semantics describe
        the \emph{difference} between the environment after and before checking
        an expression. For example, in the pair-introduction rule, $\Gamma =
        \Gamma_2 - \Gamma_1$ and $\Gamma' = \Gamma_3 - \Gamma_2$, for an
        appropriate definition of $(-)$.

@@

    \item Have \emph{one} environment where variables were \emph{tagged} as
        linear and unused, linear and used, and intuitionistic. This was
        definitely an implementation convenience so that variable binding could
        be handled uniformly for linear and intuitionistic variables and
        scoping/variable look-up could be handled implicitly thanks to the
        associative-list structure of the environment. So, the variable
        rule would most accurately look like:
        \begin{prooftree}
            \AxiomC{}
            \RightLabel{\textsc{Ty\_Var}}
            \UnaryInfC{$\Theta; \Gamma, x \overset{n}{:} t \vdash x : t ; \Gamma, x \overset{n-1}{:} t$}
        \end{prooftree}
        for $n \in \{0\,\textrm{(used)}, 1\,\textrm{(unused)},
        \omega\,\textrm{(intuitionistic)}\}$, $\omega - 1 = \omega$ and  $1 - 1 = 0$.

### 2018-05-27 write-up/related_work.tex
Finish Conclusion

\section{Matrix Expression Compilation}

% For decades, small gains in efficiency of \emph{executing} numerically
% intensive code have and continue to translate into massive savings in costs and
% time. This has spurned mountains of research into an almost bewildering variety
% of hardware, software and compilation techniques, typically entrusted with a
% (team of) domain-specific expert programmer(s) to apply in real world
% situations.
% 
% Given the rise of big data and machine learning, there is now a movement to
% improve \emph{programmer} efficiency in this domain after focusing for so long
% on \emph{computational} efficiency. Matrix expression compilers promise to
% generate code that is reasonably close to `optimal'/what a domain-expert may
% write but without the need to acquire years of experience and depth of
% knowledge (a similar trade-off to garbage-collection for memory management or
% software-transactional memory for concurrency).

@@

I believe the advantages of my approach are two-fold (1) \emph{predictable}
performance and (2) more accurate modelling of how low-level kernels handle
their resources. My confidence in the latter claim comes from finding at least
two errors in the Fortran code output by SymPy to compute a Kalman filter, one
to do with aliasing (resulting in code that was not legal Fortran) that would
not have type-checked had it been translated via LT4LA as an intermediate
representation (the other error to be explained later
in~\ref{subsec:lin_dep_types}).

@@

A particularly strong advantage LT4LA has over other libraries that use lazy
evaluation is, funnily enough, linear types, more precisely and the static and
perfect information they guarantee about aliasing: with Owl, every graph-node
has only one node in or out; with Eigen, every ``assignment'' has a
\texttt{noalias} annotation.

@@

Futhark~\cite{futhark} is a second-order (meaning it supports functions such as
map and fold/reduce) array combinator (meaning array operations can be fused
into streams to reduce temporaries) language designed for efficient parallel
compilation. It supports ML-style modules, loops, limited parametric
polymorphism, size types and uniqueness types. Intuitively, where linearity
provides a local guarantee of ``this value is not aliased in this scope'',
uniqueness types provide a global guarantee of ``this value is not aliased
anywhere''. Its combinators are more expressive than typical linear algebra
library kernels so encourages shorter, more declarative linear algebra code.

@@

Rust~\cite{rust} is a (relatively) new systems programming language aiming to
bring the last two decades of programming language research to the masses in a
usable and friendly manner. Its \emph{borrow-checker} is the feature most
relevant to this thesis because it statically attempts to prevent many
resource-related bugs. Although there are a few linear algebra libraries for
Rust under development, careful use of its macro system and borrow-checker
could make it the safest and easiest to use language for linear algebra
projects to come. Its struggle is more likely to be against the inertia of the
large amounts of C++/Fortran code already out there rather than its usability
or benefits.

@@

Dimension mismatches are seen as an irritating but small inconvenience when
writing linear algebra code. However, the second error I found in the Fortran
code output by SymPy to compute a Kalman filter was a dimension/transposition
error. Although we would not need full dependent types to solve dimension
mismatches (symbolic size types would be sufficient), managing properties about
matrices could be done at the type-level in a dependently typed setting.

>>>>

\section{Matrix Expression Compilation}\label{sec:mat_exp_comp}

@@

I believe the advantages of my approach are two-fold (1) \emph{predictable}
performance (as defined in Chapter~\ref{chap:eval}) and (2) more accurate
modelling of how low-level kernels handle their resources. My confidence in the
latter claim comes from finding errors in Fortran code output by SymPy to
compute a Kalman filter (as described in Section~\ref{sec:qual_benefits});
errors that would not have passed type-checking had it been translated via
LT4LA as an intermediate representation.

@@

A particularly strong advantage LT4LA has over other libraries that use lazy
evaluation is, funnily enough, linear types; more precisely, it is the static
and perfect information they guarantee about aliasing: with Owl, every
graph-node has only one node in or out; with Eigen, every ``assignment'' has a
\texttt{noalias} annotation, without the danger of getting it wrong and
invoking undefined behaviour.

@@

Futhark~\cite{futhark} is a second-order (meaning it supports functions such as
map and fold/reduce) array combinator (meaning array operations can be fused
into streams to reduce temporaries) language designed for efficient parallel
compilation. It supports ML-style modules, loops, limited parametric
polymorphism, size types and uniqueness types (an idea closely related to
linear types). Its combinators are more expressive than typical linear-algebra
library kernels so encourage shorter, more declarative linear-algebra code.

@@

Rust~\cite{rust} is a (relatively) new systems programming language aiming to
bring the last two decades of programming language research to the masses in a
usable and friendly manner. Its \emph{borrow-checker} is the feature most
relevant to this thesis because it statically attempts to prevent many
resource-related bugs. Although there are a few linear-algebra libraries for
Rust under development, careful use of its macro system and borrow-checker
could make it the safest and easiest to use language for linear-algebra
projects to come. Its struggle is more likely to be against the inertia of the
large amounts of C++/Fortran code already out there rather than its usability
or benefits. Given this inertia, I believe there is value in taking the DSL
library approach of LT4LA which will work with existing systems.

@@

Dimension mismatches are seen as an irritating but small inconvenience when
writing linear-algebra code. However, another error I found in the Fortran code
output by SymPy to compute a Kalman filter was a dimension/transposition error.
Although we would not need full dependent types to solve dimension mismatches
(symbolic size types would be sufficient), managing properties about matrices
could be done at the type-level in a dependently typed setting.
